---
title: "5: Data Exploration-KO"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
fig_width: 5
fig_height: 2.5
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Import and explore datasets in R
2. Graphically explore datasets in R
3. Apply data exploration skills to a real-world example dataset

## Opening discussion: why do we explore our data?

Why is data exploration our first step in analyzing a dataset? What information do we gain? How does data exploration aid in our decision-making for data analysis steps further down the pipeline?

## Import data and view summaries

```{r, message = FALSE}
# 1. Set up your working directory
getwd()

# 2. Load packages
library(tidyverse)

# 3. Import datasets
USGS.flow.data <- read.csv("./Data/Processed/USGS_Site02085000_Flow_Processed.csv")

#View(USGS.flow.data)
# Alternate option: click on data frame in Environment tab

colnames(USGS.flow.data)
str(USGS.flow.data)
dim(USGS.flow.data)

# Check our date column
class(USGS.flow.data$datetime)

USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, format = "%Y-%m-%d") #4 digit year, numeric months/numeric days with padded zeros #change from character to date format for object
class(USGS.flow.data$datetime) #check the class of datetime

```


## Visualization for Data Exploration

Although the `summary()` function is helpful in getting an idea of the spread of values in a numeric dataset, it can be useful to create visual representations of the data to help form hypotheses and direct downstream data analysis. Below is a summary of the useful types of graphs for data exploration.

Note: each of these approaches utilize the package "ggplot2". We will be covering the syntax of ggplot in a later lesson, but for now you should familiarize yourself with the functionality of what each command is doing.

### Bar Chart (function: geom_bar)

Visualize count data for categorical variables. 

```{r, fig.height = 3, fig.width = 4}
ggplot(USGS.flow.data, aes(x = discharge.mean.approval)) +
  geom_bar() 

#KO: grammar of graphics #data for plot #aes=Aesthetics #taking discharge.mean.approval attribute to break down observations to get portions of each  

#KO: made a bar plot, which is useful for categorical data

ggplot(na.omit(USGS.flow.data), aes(x = discharge.mean.approval)) +
  geom_bar() #removed NAs

```

### Histogram (function: geom_histogram)

Visualize distributions of values for continuous numerical variables. What is happening in each line of code? Insert a comment above each line.

```{r, fig.height = 3, fig.width = 4}
#KO: counts by value intervals, with numeric columns
#x-axis is a number not word

#specify number of bins between 5-20, too few miss the trends, too many only show the noise rather than the signal
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean)) 

#changed the width and to 10 units
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), binwidth = 10)

#can highlight something(ex:geom_histogram) and hit F1 for help to learn more
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = discharge.mean), bins = 5)

#force x-axis to go from 0-500 with scale command to zoom in to hist
ggplot(USGS.flow.data, aes(x = discharge.mean)) +
  geom_histogram(binwidth = 10) + 
  scale_x_continuous(limits = c(0, 500)) 
  
#gage.ht to show peak, and find a finer pattern in the data
ggplot(USGS.flow.data) +
  geom_histogram(aes(x = gage.height.mean))

```
### Frequency line graph (function: geom_freqpoly)

An alternate to a histogram is a frequency polygon graph (distributions of values for continuous numerical variables). Instead of displaying bars,  counts of continuous variables are displayed as lines. This is advantageous if you want to display multiple variables or categories of variables at once.

```{r, fig.height = 3, fig.width = 4}
#
ggplot(USGS.flow.data) +
 # geom_histogram(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.mean), bins = 50) +
  geom_freqpoly(aes(x = gage.height.min), bins = 50, color = "red") +
  geom_freqpoly(aes(x = gage.height.max), bins = 50,  lty = 2) +
  scale_x_continuous(limits = c(0, 10)) 
#lty=line type #gage ht is what value to plot

#color values with categorical value, and adding legend, gage ht mean approval assigns red to a records, and green to p records, can set color to different categorical value
ggplot(USGS.flow.data) +
  geom_freqpoly(aes(x = gage.height.mean, color = gage.height.mean.approval), bins = 50) +
  scale_x_continuous(limits = c(0, 10)) +
  theme(legend.position = "top")

```
### Box-and-whisker plots (function: geom_boxplot, geom_violin)

A box-and-whisker plot is yet another alternative to histograms (distributions of values for continuous numerical variables). These plots consist of: 

* A box from the 25th to the 75th percentile of the data, called the interquartile range (IQR).

* A bold line inside the box representing the median value of the data. Whether the median is in the center or off to one side of the IQR will give you an idea about the skewness of your data.

* A line outside of the box representing values falling within 1.5 times the IQR. 

* Points representing outliers, values that fall outside 1.5 times the IQR. 

An alternate option is a violin plot, which displays density distributions, somewhat like a hybrid of the box-and-whiskers and the frequency polygon plot.

```{r, fig.height = 3, fig.width = 4}
#displaying a numeric value, and gage ht mean approval parses into two data sets; 

#showing the median value with horizontal line in box, box shows quantiles with 50% of values in box 25-75% + outliers; many more outliers on left plot

#looking at categorical distr

ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean.approval, y = gage.height.mean))

#set the y to discharge mean, taking slices of data and taking the distribution

#group and cut width commands slice data

#showing the distr of values in the gage ht data with variations, two numeric categories

#like histograms by showing counts of slices, with tight distr and broader IQR to the right

ggplot(USGS.flow.data) +
  geom_boxplot(aes(x = gage.height.mean, y = discharge.mean, group = cut_width(gage.height.mean, 1)))

#looking at range + distr and showing the width meaning how many data points,
ggplot(USGS.flow.data) +
  geom_violin(aes(x = gage.height.mean.approval, y = gage.height.mean), 
              draw_quantiles = c(0.25, 0.5, 0.75))
```

### Scatterplot (function: geom_point)
Visualize relationships between continuous numerical variables.

```{r, fig.height = 3, fig.width = 4}
#tells us the relations among variables
#with more volume, gage ht doesn't increase, wider stream head=less volume
#trend with lots of data in lower left corner
ggplot(USGS.flow.data)+
  geom_point(aes(x = discharge.mean, y = gage.height.mean))

#hybrid btwn cat and continuous, showing mean discharge at diff times; 
#big break in data for site and some flood events w/ top outliers
#time series good wit line plot
ggplot(USGS.flow.data) +
  geom_point(aes(x = datetime, y = discharge.mean))
```

Question: under what circumstances would it be beneficial to use each of these graph types (bar plot, histogram, frequency polygon, box-and whisker, violin, scatterplot)?

> Answer: Bar plot good for categorical like polysci yes/no surveys, histogram good for showing counts of values like mean tree heights, frequency polygon good to fit monte carlo simulation, box-and whisker helpful with showing outlier events and median with numerical data sets, violin is good to show the distribution + values so for example the number of people that attended the fair at each hour of the day, and scatterplots are helpful to identify the spread of data and relationships between variables like with water coming out of a dam for energy generation.

## Ending discussion

What did you learn about the USGS discharge dataset today? What separate insights did the different graph types offer?
> Answer: The USGS data set represents stream discharge rates with values that are largely approved. Most of the time the discharge mean was below 500 and 4 for the gauge height. There were some flooding events, and a chunk of time with no data collected, and most of the time the gauge height mean was around 1.75 with over 1500 observations.  


How can multiple options for data exploration inform our understanding of our data?

> Answer: Looking at the data with multiple options helped us inspect it from different angles like holding a bowl and observing different angles to see variations in the characteristics like ht, depth, width, overall size, chips, etc.


Do you see any patterns in the USGS data for the Eno River? What might be responsible for those patterns and/or relationships?

> Answer: Yes I see there are a few flooding events that might be because of large storms with heavy rainfall, but for the most part the flow is pretty steady with a gage height right around 2 majority of the time. Perhaps a flood was so intense it disabled the gauge and that is why there was missing data. Overall, the seasonal weather variations or dam activities could account for the variations of discharge over time, but for the most part it flows consistently. 


