---
title: "5: Part 1 - Data Visualization Basics"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## Opening discussion
Effective data visualization depends on purposeful choices about graph types. The ideal graph type depends on the type of data and the message the visualizer desires to communicate. The best visualizations are clear and simple. A good resource for data visualization is [Data to Viz](https://www.data-to-viz.com/), which includes both a decision tree for visualization types and explanation pages for each type of data, including links to R resources to create them. Take a few minutes to explore this website.

## Set Up

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)
#install.packages("ggridges")
library(ggridges)

PeterPaul.chem.nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
PeterPaul.chem.nutrients.gathered <-
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv", stringsAsFactors = TRUE)
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC2021_Processed.csv", stringsAsFactors = TRUE)

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")
PeterPaul.chem.nutrients.gathered$sampledate <- as.Date(PeterPaul.chem.nutrients.gathered$sampledate, format = "%Y-%m-%d")

```

## ggplot

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms
/
Here are some commonly used layers for plotting in ggplot: 

* geom_bar #bar plot, 
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_density_ridges
* geom_point #good for lab exercises, scatterplot
* geom_errorbar
* geom_smooth #adding a trendline
* geom_line #add trendline
* geom_area
* geom_abline (plus geom_hline and geom_vline) 
    #add line with intercept a & slope b
* geom_text #add text or change defaults

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color #change the color or contour
* fill #change the fill of bar or white
* shape #decide the shape of pt
* size #size of pt using
* transparency #to points or bars

### Plotting continuous variables over time: Scatterplot and Line Plot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#window layout of ht 3, w=4 for any plot in the code chunk 
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + #x and y axis, ozone made a pt
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot) #assigning layers and then with print command for objectprint multiple plots next to each other, store as an object
#aes can be on first line or later on

# Fix this code
#changing the color
O3plot2 <- ggplot(EPAair) +
  #geom_point(aes(x = Date, y = Ozone, color = "blue")) 
  geom_point(aes(x = Date, y = Ozone), color = "blue")
print(O3plot2)
        #when color is inside () the plot adds another layer to plot with first             color available chosen (orange) with a legend
        #when outside takes it as another spec

 # Add additional variables
# How could you automatically assign a marker color to a variable? 
PMplot <- 
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year), color = Site.Name)) +  #telling r for different shapes by year, month makes columns 
                #site.name assigned by color automatically
  geom_point() #want points for variable, can increase window for better look
print(PMplot)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM2.5, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) 
print(PMplot.faceted)

#one plot with 3 rows and 4 columns, 
    #name added automatically to 12 month variable plots, 
    #can see higher and lower concentrations and numbers of observations
    #can remove sites if needed


# Filter dataset within plot building and facet by multiple variables
PMplot.faceted2 <- 
  ggplot(subset(EPAair, Site.Name == "Clemmons Middle" | Site.Name == "Leggett" |#picking 3 sites on new df, then month on x
                  Site.Name == "Bryson City"), 
         aes(x = Month, y = PM2.5)) + #Axes
  geom_point() +
  facet_grid(Site.Name ~ Year)   
print(PMplot.faceted2)
  
#still want facets with site name by year
      #setting year columns, and rows with 3 sites to see if pm changed from year to year
 #matrix look with same shapes from being in diff facets

 # Plot true time series with geom_line
PMplot.line <- 
  #line plot to connect points,
  ggplot(subset(EPAair, Site.Name == "Leggett"), 
         aes(x = Date, y = PM2.5)) + # connecting order obsv on x axis, not month
  geom_line() #making a time series plot, and picked one site

print(PMplot.line)


```


### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) + #two variables and relationship btwn them 
  geom_point()
print(lightvsDO)

# Adjust axes to look at where the points are concentrated, by omitting outliers, zooming in
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) + #excluding three points
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
#Looking at relationship btwn T and depth
#want zero to be on top and high numbers at bottom, use scale y reverse
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)



NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() + # looking at relationship btwn x = tp_ug, y = tn_ug, and categorizing by depth with color
  geom_smooth(method = lm) +#adding a trendline #linear model with blue+range
  geom_abline(aes(slope = 16, intercept = 0)) #added known line w/ sl and int
print(NvsP)

```

### Plotting continuous vs. categorical variables
A traditional way to display summary statistics of continuous variables is a bar plot with error bars. Let's explore why this might not be the most effective way to display this type of data. Navigate to the Caveats page on Data to Viz (https://www.data-to-viz.com/caveats.html) and find the page that explores barplots and error bars. 

What might be more effective ways to display the information? Navigate to the boxplots page in the Caveats section to explore further.
#Bar plots sometimes show the same display diff trends as exactly the same, even with error bars
#Hides things to the readers 
#smaller sample size, boxplot with jitter is good
#large datasets, violin plots are better with sum stats

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) + #gathered to look at all dataset nutrient concentration
    #Name and nutrient columns in PETER AND PAUL LAKES ONLY
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
  #used nutrient as category for boxplots and colors are automatically assigned
print(Nutrientplot3)  #great way to display, fill of shape in boxes, color=dots
  #color= is best to change color of boxplot pts.  

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) + #color categorizing by nutrients, and now useing dotplot(no Boxes) and categorzing with color by nutrient
  geom_dotplot(aes(color = nutrient, fill = nutrient (), binaxis = "y" 
               stackdir = "center", position = "dodge", dotsize = 2) 
print(Nutrientplot4) 
  #binaxis y = stacking on y axis 
  #filling category with color based on nutrients) 
  ##now can see something about 2nd/3rd colors,, binwidth = 1,  
  #stacking points on center position
  #dodge, no overlaying dots, wider =more observations at that value
#dotsize is chosen to be 2
#tn_ug is spread out, and yellow/green are more concentrated on values closer to zero
#recognize the ranges for diff nutrients vary, and could look concentrated, when really that could just be from the size of the y-axis


## Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(x = lakename, y = concentration)) + #nutrients categorized by color
  geom_violin(aes(color = nutrient)) #densities reflected by degree of width
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "darkred") + 
    #freqency plot with specified colors
    #need a legend with colors
    #need x-axis legend
  #inefficient
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "blue") +
  geom_freqpoly(aes(x = no23), color = "royalblue") +
  geom_freqpoly(aes(x = po4), color = "red") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.chem.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
    #gathered data with nutirients with once concen, 
    #add one geom, and categorize by color inside the aeshtetics 
    #much cleaner 
print(Nutrientplot7)

# Frequency polygons have the risk of becoming spaghetti plots. 
# See <https://www.data-to-viz.com/caveat/spaghetti.html> for more info.

# Ridgeline plot
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients.gathered, aes(y = nutrient, x = concentration)) +
  geom_density_ridges(aes(fill = lakename), alpha = 0.5) 
    #shows density of data over time, with nutrients /concentration
    #used gathered dataset
    #categorize by nutrient, fill below the line of the curves 
    #each nutrient has it's own line
print(Nutrientplot6)


```
