---
title: "6: Part 2 - Generalized Linear Models"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (ANOVA) to real datasets
3. Interpret and report the results of ANOVA in publication-style formats
>regression on categorical variables, coeff are constant to add to response
>intercept term could be the mean, and ex: shop downtown with lots of ppl vs strip mall; x+1 for mall to get mean on par with downtown

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
#install.packages("agricolae")
library(agricolae) 
#install.packages(htmltools)
#you may need to install/update package "htmltools" before installing and loading agricolae

Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is used when you have one categorical explanatory variable with more than two categories. In R, we can run the model with the function `lm` or `aov`, the latter of which will allow us to run post-hoc tests to determine pairwise differences.

Does total litter biomass differ among sites at the Niwot Ridge LTER?
```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))

summary(Litter.Totals)
#check all value of plotID
summary(Litter.Totals$plotID)

# Test for normality. 
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
# shapiro.test() performs the Shapiro-Wilk test of the null that data follows a normal distribution

shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])
#shapiro test: work with what we have, ideally want normal distr to apply anova method


#results: reject null in all except two: NIWO_057 and NIWO_046
qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass)

# Test for equal variance
# ANOVA is robust against departures from equal variance.
# bartlett.test() performs Bartlett's test of the null that the variances in each of the groups (samples) are the same.
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)

#results: reject null i.e. variances are not equal

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)

#results: reject null hypothesis i.e. difference between a pair of group means is statiscally significant

# Format ANOVA as lm
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)



# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
plot(Litter.Totals.anova2)
  #when lookingat results, the means migh tbe diff, but won't be told which are the same? use post-hoc test

# Post-hoc test
# TukeyHSD() computes Tukey Honest Significant Differences
TukeyHSD(Litter.Totals.anova) 

# Extract groupings for pairwise relationships
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE) #compare means on 2x2 basis
Litter.Totals.groups

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot) #clusters are comparable, if a letter shows up without any other letters, then it is diffe from all other letters
```
>HSD test has letter outputs with each letter corr with diff groups, with groups with abc the levels are one letter that isn't repeated, helps with identifying outliers, 
>0.57 mean is same as 067, the anova telss if they have the same mean 041-047, don't mena anything as a group
>the anova should be run if all variables having a normal distr, if more observations, then more degrees of freedom
>we needed to enter the research question, work with what we have
>ideally run Shapiro test, and make sure are on normal distr
>Shapiro test didn't have enough observations bc not enough data
>the first level will get letter a, default is letter abcd
>A site with 4 or 5 letters could be associated with different groups

What information does the output give us? How might we report this information in a report?

> 

## TWO-WAY ANOVA
### Main effects
A two-way ANOVA allows us to examine the effects of two categorical explanatory variables on a continuous response variable. What if we wanted to know how litter masses varied across functional groups and land cover class? 
>now looking at two categorical variables

```{r}
# Format as aov
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)

# Format as lm
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way)

```

### Interaction effects
We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an interaction effects two-way ANOVA, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.

```{r}
# Format as aov
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way3)


# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.
```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3) #checking for interraction

Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass)) #will show letters of two explanatory variables and making means diff across diff levels

#Litter.test <- cbind(Litter,Litter.interaction)

Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction)

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups

# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = nlcdClass, color = functionalGroup)) +
  geom_boxplot()
print(Litter.anova.plot)

```

>Questions: Monday in lab
>correlation, strength of relationship and type of correlation +/-, close to 0 is weak corr, + = increase or decrease together, - = opposite signs
>Auto-corr: how series is related to yt to yt-1, lagged observations of same variable, take column from data frame, then lag column by 1, again new column to lag by 2, and compare to og column
>partial auto-corr: relationship between y and yt-2, gives the relationship when the influence of intermediate operations, time dependence of 
>why take trend out? make a good representation of it for future projections, don't remove trend, but rather identify it; can add trend back later with models that can't take a trend in the beginning
>Identify trend with seasonal component, need a model that can represent the seasonality
>whatever happened in the past will happen in the future, good for electricity use and prices, making a decision with a fleet of power generation with supplu demand, to schedule power plants
