---
title: "Assignment 3: Data Exploration"
author: "Katherine Owens, Section #0"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**

```{r, wd and packages message = FALSE, stringsAsFactors = TRUE}
getwd()

#"C:/Users/Katherine/Documents/872-Data Analytics/Environmental_Data_Analytics_2022/Assignments"

library(dplyr)
library(ggplot2)
library(tidyverse)
```

```{r,  import datasets message = FALSE}
Neonics_data <- read.csv('../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv')

Litter_data<- read.csv('../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv')
```


## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotinoids can be toxic to sap-feeding insects like aphids and bees and cause paralysis and death of insects.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: When leaves and debris fall to the ground and break down the CO2 absorbed by photosynthesis is in turn released back into the atmosphere as a part of global seasonal variations. Also, the more deforestation there is the more woody debris that is generated from the cuttings leading to more debris that can be transported and possibly inhibit growth in other ecosystems. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Dry weight mass data of litterfall and woody debris was collected from litter traps (both elevsted and on the ground) around different plants depending on plant function such as leaves, twigs, flowers, etc.
* Collection devices included 0.5m^2square mesh ‘baskets’ elevated
~80cm above the ground, and traps on the ground were 3m x 0.5 m rectangular areas. 
*Individual sampling bouts were executed one time per year on 20 ground sites 40mx40m in size. Elevated baskets were sampled every 1-2 weeks in deciduous sites, and once every 1-2 months at evergreen sites. 

## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r data dims}
dim(Neonics_data) #has 4623 rows and 30 columns
dim(Litter_data) #has 188 rows and 30 columns
```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? 
> Answer: The most common effects are how the insecticide affected life or death reates, growth, population, behavior immunological, etc. What appeared the most though was an effect of mortality.

```{r}
head(Neonics_data$Effect,100)
summary(Neonics_data$Effect)
```


7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r summary of species counts}
head(Neonics_data$Species.Common.Name,10)
summary(Neonics_data$Species.Common.Name)
sort(table(Neonics_data$Species.Common.Name), decreasing = TRUE)
```

> Answer: Honey Bee=667,Parasitic Wasp=285, Buff Tailed Bumblebee=183, Carniolan Honey Bee=152, Bumble Bee=140, Italian Honeybee=113. All of these species fly and five out of six of them are types of bees, which are crucial for pollination of plants for agriculture. If all the bees start dying from insecticides, then farms won't be able to produce as much food and there will be shortages. We need safer insecticides for bees.  

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r class type}
class(Neonics_data$Conc.1..Author.)
head (Neonics_data$Conc.1..Author.,20)
attributes(Neonics_data$Conc.1..Author.)
```

> Answer: In the dataset the Conc.1..Author. has character arguments because there is no set format and therefore defaults to the character class most likely because of the presence of special characters like slashes and tildas. 



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}

```

> Answer: 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

```

> Answer: 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: 