---
title: 'Assignment 9: Spatial Analysis in R'
author: "Katherine Owens"
output:
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

### Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single HTML file.
6. After Knitting, please submit the completed exercise (PDF file) in Sakai. Please add your last name into the file name (e.g., "Fay_A10_SpatialAnalysis.pdf") prior to submission.

## DATA WRANGLING

### Set up your session

1. Check your working directory
2. Import libraries: tidyverse, sf, leaflet, and mapview

```{r Prepare workspace}
#1.
setwd("C:/Users/Katherine/Documents/872-Data Analytics/Environmental_Data_Analytics_2022")
getwd()


#2. 
library(tidyverse)
library(leaflet)
library(sf)
library(mapview)
library(lubridate)

```

### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3. Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4. Reveal the dataset's coordinate reference system
5. Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties
counties_sf <- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% 
  filter(STATEFP == 31) #Filter for just Nebraska Counties

#4. Reveal the CRS of the counties features
st_crs(counties_sf)$epsg #crs=4269

#5. Plot the data  
ggplot(counties_sf) +
  geom_sf()

mapview(counties_sf)

```

6. What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (Or, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: look the EPSG code on <https://spatialreference.org>) 

> ANSWER: EPSG code = 4269;  ####################### correct? ##############
> CRS = projected (NO) b/c a spheroid, this CRS uses degrees = angular; 
> and datum = North_American_Datum_1983     

>>From: https://spatialreference.org/ref/epsg/4269/html/

        

### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7. Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe.

8. Display the column names of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe
StrGageLoc <- read.csv('./Data/Raw/NWIS_SiteInfo_NE_RAW.csv', stringsAsFactors = TRUE)
class(StrGageLoc) #df, not sf-df

#8. Reveal the names of the columns
names(StrGageLoc)

```

9. What columns in the dataset contain the x and y coordinate values, respectively?  
> ANSWER: "dec_lat_va" & "dec_long_va" 



### Convert the dataframe to a spatial features ("sf") dataframe

10. Convert the dataframe to an sf dataframe.
 * Note: These data use the same coordinate reference system as the counties dataset

11. Display the column names of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object
StrGageLoc.NE.sf <- st_as_sf(StrGageLoc, coords = c('dec_long_va','dec_lat_va'),
           crs=4269)
class(StrGageLoc.NE.sf)

#11. Re-examine the column names
names(StrGageLoc)
names(StrGageLoc.NE.sf)
```

12. What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: 
> New Fields:  "dec_coord_datum_cd", "geometry"  
> Disappeared Fields: "dec_lat_va", "dec_long_va", "dec_coord_datum_cd"



### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
 * Be sure the datasets are displayed in different colors
 * Title your plot "NWIS Gage Locations in Nebraska"
 * Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features

#plot counties
ggplot() +
  geom_sf(data=counties_sf,
          aes(fill='Population'))

#plot gage locations
ggplot(StrGageLoc.NE.sf) +
  geom_sf()

#overlay map
county.gage.overlay1 <- ggplot()+
  geom_sf(data = counties_sf, color='purple', size=1) +
  geom_sf(data = StrGageLoc.NE.sf, color = 'darkgreen', size=1) +
    labs(title = "NWIS Gage Locations in Nebraska", subtitle = "Katie Owens")
county.gage.overlay1
 
```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe.
15. Show the column names.
16. Join our site information (already imported above) to these gage height data.
 * The `site_no` and `station_nm` can both/either serve as joining attributes.
 * Construct this join so that the result only includes spatial features where both tables have data.
17. Show the column names in this resulting spatial features object
18. Show the dimensions of the resulting joined dataframe

```{r}
#14. Read the site flow data into a data frame
gage.hts <- read.csv('./Data/Raw/NWIS_SiteFlowData_NE_RAW.csv', 
                    stringsAsFactors = TRUE)
#gage.hts$date <- as.Date(gage.hts$date , format = "%y-%m-%d") #fix date column
#unneccessary step^ that was removed later on.

class(gage.hts) #confirming is a df, yes!

#15. Show the column names
names(StrGageLoc)
names(gage.hts)

#16. Join location data to it
counties.gage_join <-  merge(x = gage.hts,
                           y = StrGageLoc.NE.sf, 
                           by= "site_no") 

#17. Show the column names of the joined dataset
names(counties.gage_join)

#18. Show the dimensions of this joined dataset
dim(counties.gage_join)

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.
19. Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
 * Show the magnitude of gage height by color, shape, other visualization technique.

```{r}
############don't know what I am doing wrong....Q19 says needs geometry aesthetics

#change join to sf
joined_sf <-  st_as_sf(counties.gage_join)
#(coords = c('dec_long_va','dec_lat_va'),
#           crs=4269)       #didn't need after all
class(joined_sf)

#19. Map the points, sized by gage height

ggplot() +
  geom_sf(data=counties_sf, color= "green") + #need to use sf's w/ geometry
  geom_sf(data=joined_sf, aes(color=gage_ht), alpha=0.3) +
  scale_color_gradient(low="blue", high="red")

#had to remove pipes and run st_as_sf directly on the dataframes referenced, to make it recognize the present geometry columns
#the pipes messed up the interpretation/reading/transforming df's
#use the least amount of pipes as possible
#pipes are useful if doing lots of changes on the same layer (ex: mutating multiple columns in a dataframe, if changing just one column then it's good to do it directly on the object)

```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r}
#20 Transform the counties and gage location datasets to UTM Zone 14
counties_UTM <- st_transform(counties_sf, c=32614)
class(counties_UTM)

gage.locations_UTM <- st_transform(joined_sf, c=32614)
class(gage.locations_UTM)

#21 Plot the data
ggplot(counties_UTM) + #looks the same as in #13
  geom_sf()
#map a second way
mapview(counties_UTM)

ggplot(gage.locations_UTM) + #looks the same as in #13
  geom_sf()
#map a second way
mapview(gage.locations_UTM,zcol='coord_acy_cd')
```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Lancaster county from your county sf dataframe 
23. Select the gage sites falling `within` that county
 * Use either matrix subsetting or tidy filtering 
24. Create a plot showing:
 * all Nebraska counties,
 * the selected county, 
 * and the gage sites in that county

```{r}
#22 Select the county
lancaster_county_UTM <- counties_UTM %>% 
  filter(NAME %in% 'Lancaster')
#USED a pipe bc filter & mutate only works in pipes
#don't need c unless making a list, like filtering for two diff counties


#23 Select gages within the selected county
lancaster_co_gages <- gage.locations_UTM[lancaster_county_UTM,]
#adding comma tells it to choose data from under the filtered county, vs pulling it from all the counties (undefined columns selected)
#using the [] , the comma means choo [,]

#gage.locations_UTM[,"station_nm.y"] #gives all the options in that column
#gage.locations_UTM[1:6, "station_nm.y"]


#24 Plot
ggplot() +
  geom_sf(data = counties_UTM, color='blue', size=1) +
  geom_sf(data = lancaster_county_UTM, color='orange', size=2) +
  geom_sf(data = lancaster_co_gages, color='darkgreen')
```
